---
title: "Visualising Drug Repurposing Results"
author:
  - name: Hengxin Pan
    affiliation: Walter and Eliza Hall Institute of Medical Research
    email: pan.h@wehi.edu.au
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Visualising Drug Repurposing Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

# Overview

The directional barcode heatmap is the primary visualisation in CONCERTDR.
It shows, for the top-ranked CMap signatures, whether each gene in the query
signature is up- or down-regulated in the drug-perturbed profile — providing
intuition for *why* a drug scored well.

The workflow has two steps:

1. **`extract_signature_zscores()`** — reads z-scores from the reference
   matrix (or GCTX file) for the selected perturbations and returns an R list.
   This is the computationally expensive step; save the output and reuse it.
2. **`plot_signature_direction_tile_barcode()`** — takes the precomputed list
   and renders the heatmap. Can be called repeatedly with different clustering
   or display options without re-reading any files.

```{r load}
library(CONCERTDR)
```

# Preparing the Input Data

## Scoring results

```{r score}
sig_file     <- system.file("extdata", "example_signature.txt",
                             package = "CONCERTDR")
ref_file     <- system.file("extdata", "example_reference_df.csv",
                             package = "CONCERTDR")
reference_df <- read.csv(ref_file, row.names = 1, check.names = FALSE)
reference_df$gene_symbol <- rownames(reference_df)

scoring_results <- process_signature_with_df(
  signature_file = sig_file,
  reference_df   = reference_df,
  methods        = "ks",
  permutations   = 100,
  save_files     = FALSE
)
results_df <- scoring_results$results$ks
head(results_df[order(results_df$Score), ])
```

# Step 1 — Extracting Z-scores

`extract_signature_zscores()` handles three modes:

| Mode | When to use |
|------|------------|
| `reference_df =` (in-memory matrix) | Bundled data or pre-built reference |
| `gctx_file =` | Full CMap GCTX file on disk |
| Auto-resolve via `options()` | File paths set globally at script top |

## Using an in-memory reference matrix

```{r extract-zscores}
z <- extract_signature_zscores(
  results_df     = results_df,
  signature_file = sig_file,
  reference_df   = reference_df,
  max_genes      = 20,    # use all 20 genes in the example
  max_perts      = 10,    # show all 10 example signatures
  verbose        = TRUE
)
```

## Inspecting the returned object

```{r inspect-z}
names(z)
```

| Element | Description |
|---------|-------------|
| `z_plot` | Matrix — rows = perturbations, cols = genes (down→up order) |
| `ordered_genes` | Character vector of genes in barcode order |
| `logfc_map` | Named numeric vector of signature log2FC values |
| `sig_ids` | CMap signature IDs used |
| `sig_labels` | Human-readable row labels (`drug | dose | time | cell`) |

```{r inspect-dimensions}
cat("Matrix dimensions:", nrow(z$z_plot), "perturbations ×",
    ncol(z$z_plot), "genes\n")

# Genes are ordered: down-regulated (left) → up-regulated (right)
cat("First 5 genes (down-regulated in disease):\n",
    head(z$ordered_genes, 5), "\n")
cat("Last 5 genes (up-regulated in disease):\n",
    tail(z$ordered_genes, 5), "\n")

# Row labels combine drug name, dose, time, and cell line
head(z$sig_labels, 4)
```

## Gene ordering rationale

The columns of `z_plot` follow a fixed order: genes down-regulated in the
disease signature (sorted by log2FC, most negative first) are placed on the
left; up-regulated genes (most positive first) are on the right.

A drug that perfectly *reverses* the signature will show high z-scores (warm
colours) on the left (disease-down genes suppressed by disease → activated by
drug) and low z-scores (cool colours) on the right (disease-up genes
activated by disease → suppressed by drug).

```{r logfc-distribution}
# Confirm the ordering: negative logFC on left, positive on right
lfc_ordered <- z$logfc_map[z$ordered_genes]
cat("Range of logFC for left half (down genes):",
    round(range(lfc_ordered[seq_len(sum(lfc_ordered < 0))]), 2), "\n")
cat("Range of logFC for right half (up genes):",
    round(range(lfc_ordered[seq_len(sum(lfc_ordered > 0)) + sum(lfc_ordered < 0)]), 2), "\n")
```

## Saving the z-score matrix

Save the matrix as a TSV for downstream use or sharing:

```{r save-zscores, eval=FALSE}
z <- extract_signature_zscores(
  results_df     = results_df,
  signature_file = sig_file,
  reference_df   = reference_df,
  output_zscores = "results/barcode_zscores.tsv"
)
```

The file can be reloaded and passed back as `reference_df` in a subsequent
call to avoid re-running the extraction step.

# Step 2 — Plotting the Barcode Heatmap

`plot_signature_direction_tile_barcode()` requires `ComplexHeatmap` and
`circlize`. The `precomputed` argument accepts the output of
`extract_signature_zscores()` and skips all data loading.

```{r heatmap, fig.width=10, fig.height=6}
if (requireNamespace("ComplexHeatmap", quietly = TRUE) &&
    requireNamespace("circlize", quietly = TRUE)) {

  plot_signature_direction_tile_barcode(
    precomputed  = z,
    cluster_rows = FALSE,   # keep score-order rows for now
    cluster_cols = FALSE,   # keep signature gene order
    verbose      = FALSE
  )

} else {
  message("Install ComplexHeatmap and circlize for heatmap rendering:\n",
          "  BiocManager::install('ComplexHeatmap')\n",
          "  install.packages('circlize')")
}
```

## With row clustering

Clustering the perturbation rows reveals groups of drugs with similar
transcriptional responses to the query genes.

```{r heatmap-clustered, fig.width=10, fig.height=6}
if (requireNamespace("ComplexHeatmap", quietly = TRUE) &&
    requireNamespace("circlize", quietly = TRUE)) {

  plot_signature_direction_tile_barcode(
    precomputed         = z,
    cluster_rows        = TRUE,
    cluster_method      = "complete",
    show_row_dendrogram = TRUE,
    cluster_cols        = FALSE,
    verbose             = FALSE
  )
}
```

## Saving to PNG

```{r heatmap-png, eval=FALSE}
plot_signature_direction_tile_barcode(
  precomputed  = z,
  cluster_rows = TRUE,
  save_png     = TRUE,
  output_png   = "results/barcode_heatmap.png",
  width        = 14,     # inches; NULL = auto-compute from data size
  height       = 8,
  dpi          = 150
)
```

# Colour Scheme

Two independent colour scales are used, chosen to be visually separable:

| Scale | Colours | Meaning |
|-------|---------|---------|
| Heatmap z-scores | `#3B4CC0` (blue) → white → `#B40426` (red) | Coolwarm; cold = repressed, warm = activated |
| Log2FC annotation bar | `#01665E` (teal) → white → `#8C510A` (brown) | BrBG; teal = down-regulated in disease, brown = up-regulated |

The annotation bar at the top of the heatmap shows the log2FC direction of
each gene in the query signature, making it immediately clear which genes are
expected to be down (left) vs. up (right) in the disease context.

# Filtering to a Single Drug

When `annotate_drug_results()` has been run, the `tech_view_all` table
contains multiple rows per drug (one per CMap context). Pass this table as
`results_df` and use `selected_drug` to display only the contexts for one
drug of interest:

```{r single-drug, eval=FALSE}
# With full CMap data after annotate_drug_results():
z_imatinib <- extract_signature_zscores(
  results_df        = views$tech_view_all,
  signature_file    = sig_file,
  selected_drug     = "imatinib",
  selected_drug_col = "display_name",
  max_perts         = 30
)

plot_signature_direction_tile_barcode(
  precomputed = z_imatinib,
  cluster_rows = TRUE,
  save_png     = FALSE
)
```

# Using the Full CMap GCTX File

When the GCTX file is available, `extract_signature_zscores()` reads z-scores
directly from the file. File paths are resolved in priority order:

1. Explicit `gctx_file` argument
2. `options(CONCERTDR.gctx_file = ...)`
3. `Sys.getenv("CONCERTDR_GCTX_FILE")`
4. `data_dir/level5_beta_all_n1201944x12328.gctx`

```{r gctx-workflow, eval=FALSE}
# Set global paths once
options(
  CONCERTDR.gctx_file     = "/data/cmap/level5_beta_all_n1201944x12328.gctx",
  CONCERTDR.geneinfo_file = "/data/cmap/geneinfo_beta.txt",
  CONCERTDR.siginfo_file  = "/data/cmap/siginfo_beta.txt"
)

# extract_signature_zscores picks up paths from options automatically
z <- extract_signature_zscores(
  results_df     = views$tech_view_all,
  signature_file = "my_signature.txt",
  max_genes      = 100,
  max_perts      = 60,
  output_zscores = "results/barcode_zscores.tsv"
)

# Plot — reuse the precomputed object for all iterations
plot_signature_direction_tile_barcode(
  precomputed  = z,
  cluster_rows = TRUE,
  save_png     = TRUE,
  output_png   = "results/barcode_heatmap.png"
)
```

# Performance Tips

- **Save `z` and reuse it.** `extract_signature_zscores()` reads from GCTX or
  the reference matrix; `plot_signature_direction_tile_barcode(precomputed = z)`
  does no I/O at all. If you want to experiment with clustering parameters or
  figure dimensions, call the plot function repeatedly with the same `z`.

- **Limit `max_genes` and `max_perts` for exploration.** The bundled example
  has 20 genes and 10 signatures. For real data, `max_genes = 100` and
  `max_perts = 60` produce readable figures; adjust with `width` and `height`
  or let the function auto-scale (`~0.22 in/gene`, `~0.28 in/perturbation`).

- **Use `use_raster = TRUE` (default).** ComplexHeatmap's raster rendering
  keeps PNG file sizes manageable when `max_perts` is large.

# Session Info

```{r session-info}
sessionInfo()
```
