---
title: "CMap Data Extraction Workflow"
author: "Hengxin Pan"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{CMap Data Extraction Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# CMap Data Extraction Workflow

This vignette walks through extracting a reference expression matrix from the
CMap L1000 database using CONCERTDR, then saving it for downstream signature
matching. All runnable code uses the small bundled example datasets; code that
requires the full CMap download is wrapped in `\donttest{}` / `eval=FALSE`.

```{r setup}
library(CONCERTDR)
```

## Overview of the Workflow

```
siginfo_beta.txt  ──► extract_cmap_parameters()  ──► config file
                                                         │
                                                         ▼
geneinfo_beta.txt ─┐                         generate_combinations_from_config()
siginfo_beta.txt  ─┤                                     │
gctx_file         ─┘                                     ▼
                                         extract_cmap_data_from_config()
                                                         │
                                                         ▼
                                              reference_df  (genes × signatures)
```

## Step 1: Inspect the Bundled Example Data

The package includes miniature versions of the CMap data files for testing:

```{r inspect-example}
# Paths to bundled example files
siginfo_ex  <- system.file("extdata", "example_siginfo.txt",     package = "CONCERTDR")
geneinfo_ex <- system.file("extdata", "example_geneinfo.txt",    package = "CONCERTDR")
ref_ex      <- system.file("extdata", "example_reference_df.csv", package = "CONCERTDR")

# Siginfo: one row per CMap signature
siginfo <- read.delim(siginfo_ex)
cat("Signatures:", nrow(siginfo), "\n")
head(siginfo[, c("sig_id", "cmap_name", "pert_idose", "pert_itime", "cell_iname")])
```

```{r inspect-geneinfo}
# Geneinfo: one row per gene
geneinfo <- read.delim(geneinfo_ex)
cat("Genes:", nrow(geneinfo), "\n")
head(geneinfo[, c("gene_id", "gene_symbol", "feature_space")])
```

```{r inspect-ref}
# Reference data frame: genes as rows, signatures as columns
ref_df <- read.csv(ref_ex, row.names = 1, check.names = FALSE)
cat("Dimensions:", nrow(ref_df), "genes x", ncol(ref_df), "signatures\n")
ref_df[1:5, 1:5]
```

## Step 2: Extract Parameters from a Siginfo File

In a real analysis, run `extract_cmap_parameters()` on the full
`siginfo_beta.txt` to discover all available time points, doses, and cell
lines, and to write a configuration file:

```{r extract-params, eval=FALSE}
# Requires the full siginfo_beta.txt downloaded from clue.io
params <- extract_cmap_parameters(
  siginfo_file = "path/to/siginfo_beta.txt",
  config_dir   = "conf",
  verbose      = TRUE
)
cat("Time points:", length(params$times), "\n")
cat("Doses:",       length(params$doses), "\n")
cat("Cell lines:",  length(params$cells), "\n")
```

With the bundled example file we can demonstrate the same function:

```{r extract-params-example}
params <- extract_cmap_parameters(
  siginfo_file = siginfo_ex,
  config_dir   = tempdir(),
  verbose      = FALSE
)
cat("Example time points:", paste(params$times, collapse = ", "), "\n")
cat("Example doses:",       paste(params$doses, collapse = ", "), "\n")
cat("Example cell lines:",  paste(params$cells, collapse = ", "), "\n")
```

## Step 3: Create a Configuration File

`create_cmap_config_template()` writes a template configuration file that you
can edit to select specific parameter combinations:

```{r create-config}
cfg_dir  <- tempdir()
cfg_file <- create_cmap_config_template(
  dest_dir      = cfg_dir,
  template_name = "my_config.conf",
  overwrite     = TRUE
)
cat("Config written to:", cfg_file, "\n")

# Inspect the [selected] section
cfg_lines <- readLines(cfg_file)
cat(tail(cfg_lines, 15), sep = "\n")
```

## Step 4: Generate Parameter Combinations

`create_console_config()` reads a siginfo file and lets the user choose
time points, doses, and cell lines; in non-interactive (batch) mode it
selects all available options automatically.
`generate_combinations_from_selections()` then expands the choices into
every combination:

```{r combinations}
# In non-interactive sessions all available options are selected automatically
selections   <- create_console_config(siginfo_ex, config_dir = tempdir(),
                                      verbose = FALSE)
combinations <- generate_combinations_from_selections(selections, verbose = FALSE)
combinations
```

## Step 5: Extract Reference Data from GCTX

`extract_cmap_data_from_config()` reads the GCTX file for every combination
and returns a single data frame (genes x all signatures).

```{r extract-gctx, eval=FALSE}
# Requires the full CMap files downloaded from clue.io
ref_df <- extract_cmap_data_from_config(
  config_file   = cfg_file,
  geneinfo_file = "path/to/geneinfo_beta.txt",
  siginfo_file  = "path/to/siginfo_beta.txt",
  gctx_file     = "path/to/level5_beta_trt_cp_n720216x12328.gctx",
  landmark      = TRUE,
  verbose       = TRUE
)
cat("Extracted:", nrow(ref_df), "genes x", ncol(ref_df), "signatures\n")

# Save for later use
write.csv(ref_df, "reference_df.csv")
```

Alternatively, filter the siginfo file first with `subset_siginfo_beta()`,
then pass it directly to `extract_cmap_data_from_siginfo()`:

```{r extract-siginfo, eval=FALSE}
# Non-interactive filtering
filtered_sig <- subset_siginfo_beta(
  "path/to/siginfo_beta.txt",
  interactive = FALSE,
  filters = list(
    pert_itime = "24 h",
    pert_idose = "10 uM",
    cell_iname = c("K562", "MCF7", "A375")
  ),
  output_file = "filtered_siginfo.txt"
)

ref_df <- extract_cmap_data_from_siginfo(
  siginfo_file  = "filtered_siginfo.txt",
  geneinfo_file = "path/to/geneinfo_beta.txt",
  gctx_file     = "path/to/level5_beta_trt_cp_n720216x12328.gctx"
)
```

## Using the Bundled Reference Data in Downstream Analysis

The bundled `example_reference_df.csv` mimics the output of Step 5 and can
be passed directly to `process_signature_with_df()` (see the
*Signature Matching* vignette):

```{r use-example-ref}
ref_df <- read.csv(ref_ex, row.names = 1, check.names = FALSE)
cat("Ready for signature matching:", nrow(ref_df), "genes x",
    ncol(ref_df), "signatures\n")
```

## Session Info

```{r session-info}
sessionInfo()
```
