---
title: "Introduction to CONCERTDR"
author: "Hengxin Pan"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to CONCERTDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# CONCERTDR: Drug Response Data Analysis Tools

## Overview

CONCERTDR provides a complete pipeline for drug repurposing analysis using
the Connectivity Map (CMap) L1000 database. The main steps are:

1. **CMap data extraction** - subset the GCTX file by time, dose, and cell
   line to obtain a reference expression matrix.
2. **Signature matching** - score an input differential-expression signature
   against the reference matrix using KS, GSEA, XCos, XSum, and Zhang methods.
3. **Result annotation** - integrate scores with compound metadata
   (MoA, targets, clinical phase) from CMap compound info files.
4. **Visualisation** - directional barcode heatmaps via ComplexHeatmap.

## Installation

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("CONCERTDR")
```

## Acquiring CMap Data

Download the following files from [clue.io](https://clue.io/) after free
registration:

| File | Description |
|------|-------------|
| `siginfo_beta.txt` | Metadata for all ~720k signatures |
| `geneinfo_beta.txt` | Gene annotations (978 landmark genes) |
| `level5_beta_trt_cp_n720216x12328.gctx` | Expression matrix (~14 GB) |
| `compoundinfo_beta.txt` | Drug metadata (MoA, targets, phase) |

## Quick-start with Bundled Example Data

The package ships with small example datasets so every function can be
explored without downloading the full CMap files. These bundled files support
signature matching and summary generation; GCTX-dependent extraction and
visualisation still require the external CMap resources.

```{r setup}
library(CONCERTDR)

# Locate bundled example files
sig_file <- system.file("extdata", "example_signature.txt",    package = "CONCERTDR")
ref_file <- system.file("extdata", "example_reference_df.csv", package = "CONCERTDR")

# Read the example reference data frame
# In a real analysis this is produced by extract_cmap_data_from_config()
ref_df <- read.csv(ref_file, row.names = 1, check.names = FALSE)
cat("Reference matrix:", nrow(ref_df), "genes x", ncol(ref_df), "signatures\n")

# Preview the query signature (Gene + log2FC columns)
sig <- read.delim(sig_file)
head(sig)
```

## Package Modules

| Module | Key functions |
|--------|---------------|
| Data extraction | `extract_cmap_parameters()`, `generate_combinations_from_config()`, `extract_cmap_data_from_config()`, `extract_cmap_data_from_siginfo()` |
| Siginfo filtering | `subset_siginfo_beta()` |
| Signature matching | `process_signature_with_df()`, `run_cmap_workflow()` |
| Result annotation | `annotate_drug_results()`, `fuzzy_drug_match()` |
| Visualisation | `extract_signature_zscores()`, `plot_signature_direction_tile_barcode()` |

See the companion vignettes for detailed walkthroughs:

- **CMap Workflow** - extracting reference data from the GCTX file
- **Signature Matching** - matching and annotating results end-to-end

## Session Info

```{r session-info}
sessionInfo()
```
