---
title: "Introduction to CONCERTDR"
author:
  - name: Hengxin Pan
    affiliation: Walter and Eliza Hall Institute of Medical Research
    email: pan.h@wehi.edu.au
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to CONCERTDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE
)
```

# Overview

**CONCERTDR** (**CO**ntext-aware **N**etwork-based **CE**llular and
**R**epurposing **T**ool for **D**rug **R**epurposing) provides a complete
computational pipeline for identifying candidate repurposed drugs by matching
disease differential-expression signatures against the Connectivity Map (CMap)
L1000 perturbation database.

The main steps of the workflow are:

1. **Filter** the CMap signature metadata (`siginfo_beta.txt`) to the
   cell lines and perturbation types relevant to your disease context.
2. **Build** a reference expression matrix from the CMap GCTX file.
3. **Score** your disease signature against the reference matrix using one or
   more connectivity scoring methods.
4. **Annotate** hits with compound metadata (mechanism of action, clinical
   phase, targets).
5. **Visualise** the top hits as a directional barcode heatmap.

All steps in this vignette use the small example dataset bundled with the
package, so no external downloads are required to run the code.

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("CONCERTDR")

# Optional but recommended for visualisation
BiocManager::install("ComplexHeatmap")
install.packages("circlize")
```

```{r load}
library(CONCERTDR)
```

# Bundled Example Data

The package ships with a self-contained toy dataset in `inst/extdata/`:

| File | Description |
|------|-------------|
| `example_signature.txt` | 20-gene disease signature with log2FC values |
| `example_reference_df.csv` | 20 genes × 10 CMap-style signatures |
| `example_siginfo.txt` | Metadata for the 10 example signatures |
| `example_geneinfo.txt` | Gene annotation for the 20 landmark genes |

Use `system.file()` to locate these files — this function looks inside the
installed package directory. **Do not** use `system.file()` for your own data
files on disk; use plain `file.path()` or absolute paths instead.

```{r bundled-files}
sig_file  <- system.file("extdata", "example_signature.txt",
                          package = "CONCERTDR")
ref_file  <- system.file("extdata", "example_reference_df.csv",
                          package = "CONCERTDR")
sig_info_file <- system.file("extdata", "example_siginfo.txt",
                              package = "CONCERTDR")
gene_info_file <- system.file("extdata", "example_geneinfo.txt",
                               package = "CONCERTDR")
```

## The query signature

A signature file is a tab-delimited text file with two required columns:

- `Gene`: HGNC gene symbols (uppercase)
- `log2FC`: log2 fold-change (positive = up-regulated, negative = down-regulated)

Rows should be sorted by `|log2FC|` descending so that the most informative
genes are used first when `max_genes` is applied in later steps.

```{r read-signature}
signature <- read.delim(sig_file)
head(signature, 10)
```

```{r signature-summary}
cat("Up-regulated genes:  ", sum(signature$log2FC > 0), "\n")
cat("Down-regulated genes:", sum(signature$log2FC < 0), "\n")
```

## The reference matrix

In a full analysis this matrix is produced by
`extract_cmap_data_from_siginfo()` (see the *Signature Matching* vignette).
Here we use the pre-computed CSV bundled with the package.

```{r read-reference}
reference_df <- read.csv(ref_file, row.names = 1, check.names = FALSE)
# The gene_symbol column is required by process_signature_with_df()
reference_df$gene_symbol <- rownames(reference_df)

cat("Reference dimensions:", nrow(reference_df), "genes ×",
    ncol(reference_df) - 1, "signatures\n")
reference_df[1:5, 1:6]
```

# A Minimal End-to-End Run

The function `process_signature_with_df()` is the central workhorse of
CONCERTDR. It accepts the signature file, the reference matrix, and a choice
of scoring methods, and returns a structured `cmap_signature_result` object.

```{r run-scoring}
results <- process_signature_with_df(
  signature_file = sig_file,
  reference_df   = reference_df,
  methods        = c("ks", "xsum"),
  topN           = 4,          # genes used by XSum; increase for real data
  permutations   = 100,        # increase to ≥1000 for publication-quality p-values
  save_files     = FALSE
)
```

## Inspecting the result object

`print()` gives a compact overview:

```{r print-results}
print(results)
```

`summary()` lists the top hits across all methods ranked by a combined score:

```{r summary-results}
summary(results, top_n = 5)
```

Individual method results are data frames inside `results$results`:

```{r inspect-ks}
ks_df <- results$results$ks
ks_df[order(ks_df$Score), ]
```

**Score interpretation:** a negative score indicates that the drug *reverses*
the disease signature (potentially therapeutic); a positive score indicates
that it *mimics* the signature (possibly aggravating).

## Quick plots

The `plot()` method supports two visualisation types out of the box.

```{r plot-scores, fig.width=7, fig.height=4}
plot(results, method = "ks", plot_type = "scores", top_n = 10)
```

```{r plot-volcano, fig.width=6, fig.height=4}
plot(results, method = "ks", plot_type = "volcano")
```

# Available Scoring Methods

CONCERTDR implements seven connectivity scoring methods internally:

| Method | Argument | Reference |
|--------|----------|-----------|
| Kolmogorov–Smirnov | `"ks"` | Lamb *et al.* 2006 |
| GSEA (weight 0) | `"gsea0"` | Subramanian *et al.* 2005 |
| GSEA (weight 1) | `"gsea1"` | Subramanian *et al.* 2005 |
| GSEA (weight 2) | `"gsea2"` | Subramanian *et al.* 2005 |
| Extreme cosine | `"xcos"` | Cheng *et al.* 2014 |
| Extreme sum | `"xsum"` | Cheng *et al.* 2014 |
| Zhang score | `"zhang"` | Zhang & Gant 2008 |

Run all methods at once by passing `methods = c("ks","xcos","xsum","gsea0","gsea1","gsea2","zhang")`.
For real data, `topN = 400` and `permutations = 1000` are reasonable starting
points; lower values are used here to keep the vignette fast.

# Setting Up for Full CMap Data

When working with the full CMap release, set global options once at the top of
your script so that all CONCERTDR functions can resolve the file paths
automatically:

```{r set-options, eval=FALSE}
data_dir <- "/path/to/your/cmap_data"

options(
  CONCERTDR.data_dir          = data_dir,
  CONCERTDR.gctx_file         = file.path(data_dir, "level5_beta_all_n1201944x12328.gctx"),
  CONCERTDR.siginfo_file      = file.path(data_dir, "siginfo_beta.txt"),
  CONCERTDR.geneinfo_file     = file.path(data_dir, "geneinfo_beta.txt"),
  CONCERTDR.compoundinfo_file = file.path(data_dir, "compoundinfo_beta.txt")
)

# Verify paths before running
stopifnot(file.exists(getOption("CONCERTDR.gctx_file")))
stopifnot(file.exists(getOption("CONCERTDR.siginfo_file")))
```

CMap data files are available at <https://clue.io/releases/data-dashboard>
after free registration.

# Further Reading

For detailed walkthroughs of each stage:

- **Signature Matching** vignette — full workflow from `subset_siginfo_beta()`
  through `process_signature_with_df()`.
- **Result Annotation** vignette — integrating CMap compound metadata with
  `annotate_drug_results()`.
- **Visualization** vignette — directional barcode heatmaps with
  `extract_signature_zscores()` and `plot_signature_direction_tile_barcode()`.

# Session Info

```{r session-info}
sessionInfo()
```
