---
title: "Annotating Drug Repurposing Results"
author:
  - name: Hengxin Pan
    affiliation: Walter and Eliza Hall Institute of Medical Research
    email: pan.h@wehi.edu.au
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Annotating Drug Repurposing Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

# Overview

After running `process_signature_with_df()`, each row of the results table
corresponds to a single CMap signature ID (a specific drug × dose × time ×
cell-line combination). To make the results interpretable for biologists, we
need to:

1. Map signature IDs back to human-readable drug names.
2. Add compound metadata: mechanism of action (MoA), targets, clinical phase.
3. Summarise across contexts (same drug tested in multiple cell lines/doses).
4. Identify which hits are drugs (small molecules) vs. gene-knockdown (shRNA/CRISPR).

`annotate_drug_results()` performs all of these steps in a single call and
returns four analysis-ready tables.

```{r load}
library(CONCERTDR)
```

# Input Data

## Scoring results

We use the bundled example data to generate KS scores:

```{r run-scoring}
sig_file     <- system.file("extdata", "example_signature.txt",
                             package = "CONCERTDR")
ref_file     <- system.file("extdata", "example_reference_df.csv",
                             package = "CONCERTDR")
reference_df <- read.csv(ref_file, row.names = 1, check.names = FALSE)
reference_df$gene_symbol <- rownames(reference_df)

scoring_results <- process_signature_with_df(
  signature_file = sig_file,
  reference_df   = reference_df,
  methods        = "ks",
  permutations   = 100,
  save_files     = FALSE
)

results_df <- scoring_results$results$ks
head(results_df[order(results_df$Score), ])
```

## Required metadata files

`annotate_drug_results()` requires two metadata files (or data frames):

| Input | Source | Key columns |
|-------|--------|-------------|
| `sig_info_file` | `siginfo_beta.txt` from clue.io, or the data frame from `subset_siginfo_beta()` | `sig_id`, `pert_type`, `pert_id` |
| `comp_info_file` | `compoundinfo_beta.txt` from clue.io | `pert_id`, `cmap_name`, `moa`, `target`, `phase` |

The bundled example siginfo is used here. For a real analysis, pass
`sig_info_file = filtered_siginfo` (the data frame returned by
`subset_siginfo_beta()`) directly — there is no need to write it to disk.

```{r load-metadata}
sig_info_file <- system.file("extdata", "example_siginfo.txt",
                              package = "CONCERTDR")
siginfo       <- read.delim(sig_info_file, stringsAsFactors = FALSE)
head(siginfo)
```

For demonstration purposes we build a small compoundinfo data frame that
mimics the structure of `compoundinfo_beta.txt`:

```{r build-compinfo}
compinfo <- data.frame(
  pert_id           = c("BRD-K00001", "BRD-K00002", "BRD-K00003",
                         "BRD-K00004", "BRD-K00005"),
  pert_name         = c("imatinib",    "dasatinib",   "doxorubicin",
                         "methotrexate", "vincristine"),
  cmap_name         = c("imatinib",    "dasatinib",   "doxorubicin",
                         "methotrexate", "vincristine"),
  target            = c("BCR-ABL|KIT", "BCR-ABL|SRC|LCK",
                         "TOP2A",        "DHFR",   "TUBB"),
  moa               = c("kinase inhibitor", "kinase inhibitor",
                         "DNA synthesis inhibitor",
                         "antimetabolite", "microtubule inhibitor"),
  phase             = c("Launched", "Launched", "Launched",
                         "Launched", "Launched"),
  stringsAsFactors  = FALSE
)

# Patch the bundled siginfo to include matching pert_id values
siginfo$pert_id <- compinfo$pert_id[
  match(siginfo$cmap_name, compinfo$cmap_name)
]
head(siginfo[, c("sig_id", "cmap_name", "pert_id", "pert_type")])
```

# Running `annotate_drug_results()`

```{r annotate}
views <- annotate_drug_results(
  results_df     = results_df,
  sig_info_file  = siginfo,     # data.frame, not a file path
  comp_info_file = compinfo,
  write_outputs  = FALSE,
  verbose        = TRUE
)
```

The function returns a named list with five elements:

```{r view-names}
names(views)
```

# The Four Output Tables

## `tech_view_all` — Full technical table

Contains one row per input signature ID with all integrated fields. This is
the master table used by downstream visualisation functions.

```{r tech-view}
tv <- views$tech_view_all
# Key columns for drug repurposing interpretation
tv[order(tv$Score),
   c("compound", "display_name", "Score", "pAdjValue",
     "effect_direction", "cell_line", "time_h", "dose_uM",
     "moa", "target")]
```

### Effect direction

```{r effect-direction}
table(tv$effect_direction)
```

Negative scores indicate that the drug *reverses* the disease signature —
this is the standard criterion for candidate therapeutic repurposing.

### Perturbation kind

The `pert_kind` column categorises each hit as a Drug, Gene (knockdown or
overexpression), or Control:

```{r pert-kind}
table(tv$pert_kind, tv$pert_type)
```

## `wetlab_drug_view` — Drug-focused summary

De-duplicated to one row per drug (lowest score across all contexts).
Intended for biologists who want a concise hit list for follow-up experiments.

```{r wetlab-drug}
views$wetlab_drug_view[,
  c("display_name", "Score", "effect_direction",
    "target", "moa", "phase")]
```

### Including dose in the drug view

By default `dose_uM` is omitted from the wetlab drug view because the table
is de-duplicated across contexts. Set `keep_dose_in_drug_view = TRUE` to
retain it:

```{r wetlab-drug-dose}
views_with_dose <- annotate_drug_results(
  results_df             = results_df,
  sig_info_file          = siginfo,
  comp_info_file         = compinfo,
  keep_dose_in_drug_view = TRUE,
  write_outputs          = FALSE,
  verbose                = FALSE
)
views_with_dose$wetlab_drug_view[, c("display_name", "Score", "dose_uM")]
```

## `wetlab_gene_view` — Gene perturbation hits

Contains only `trt_sh`, `trt_xpr`, and `trt_oe` perturbations. Empty in this
example (all perturbations are drugs), but populated in a full CMap analysis
that includes gene knockdown screens.

```{r wetlab-gene}
nrow(views$wetlab_gene_view)
# In a full analysis this would list top shRNA/CRISPR hits:
# views$wetlab_gene_view[, c("pert_name", "mode", "Score", "cell_line")]
```

## `drug_context_summary` — Context breadth table

One row per drug, with counts showing in how many distinct contexts
(cell lines, doses, time points) the drug appeared as a hit. Drugs that
score well across many contexts are generally more compelling candidates.

```{r drug-context}
views$drug_context_summary[,
  c("display_name", "best_score", "n_contexts",
    "n_cell_lines", "moa_status")]
```

`moa_status` is `"Known"` when a MoA entry exists in the compound info and
`"Unknown"` otherwise — useful for identifying drugs with unexplored mechanisms.

# Writing Outputs to Disk

Set `write_outputs = TRUE` and supply `output_dir` to save all four tables as
tab-separated files:

```{r write-outputs, eval=FALSE}
views_saved <- annotate_drug_results(
  results_df     = results_df,
  sig_info_file  = "path/to/siginfo_beta.txt",
  comp_info_file = "path/to/compoundinfo_beta.txt",
  output_dir     = "results",
  write_outputs  = TRUE
)
# Written files:
# results/wetlab_drug_view.tsv
# results/wetlab_gene_view.tsv
# results/tech_view_all.tsv
# results/drug_context_summary.tsv
```

The paths of written files are returned in `views$output_files`:

```{r show-output-files, eval=FALSE}
views_saved$output_files
```

# Compound ID Utilities

## `extract_compound_id()`

CMap signature IDs use a structured colon-delimited format:

```
LIBRARY_CELL_TIME : BRD_ID : DOSE : TIME
```

`extract_compound_id()` extracts any field by splitting on `:` or `_`, or by
a custom regular expression.

```{r extract-compound-id}
example_ids <- c(
  "CVD001_K562_6H:BRD-K52492843-001-01-8:10:6",
  "CVD001_HL60_24H:BRD-A00619745-001-03-9:1:24"
)

# Extract the BRD identifier (field 2)
extract_compound_id(example_ids, method = "split_colon", part_index = 2)

# Extract using a regular expression
extract_compound_id(example_ids, method = "regex",
                    regex_pattern = "BRD-[A-Z0-9-]+")

# Extract the cell line from the first field
extract_compound_id(
  vapply(strsplit(example_ids, ":"), `[[`, character(1), 1),
  method = "split_underscore",
  part_index = 2
)
```

## `fuzzy_drug_match()`

When reconciling drug names across databases (e.g., CMap vs. a clinical trial
registry), name spelling may differ. `fuzzy_drug_match()` uses string
similarity to find the closest matches.

```{r fuzzy-match}
if (requireNamespace("RecordLinkage", quietly = TRUE)) {
  query_names <- c("imatinb", "doxorubicn", "vincristin")  # deliberate typos
  ref_names   <- c("imatinib", "dasatinib", "doxorubicin",
                   "methotrexate", "vincristine")

  matches <- fuzzy_drug_match(
    query_names     = query_names,
    reference_names = ref_names,
    method          = "levenshtein",
    threshold       = 70
  )
  matches
} else {
  message("Install RecordLinkage for fuzzy name matching: ",
          "install.packages('RecordLinkage')")
}
```

# Full CMap Workflow (Reference)

For completeness, here is the full workflow using the real CMap files:

```{r full-workflow, eval=FALSE}
# 1. Set paths once
options(
  CONCERTDR.gctx_file         = "/data/cmap/level5_beta_all_n1201944x12328.gctx",
  CONCERTDR.siginfo_file      = "/data/cmap/siginfo_beta.txt",
  CONCERTDR.geneinfo_file     = "/data/cmap/geneinfo_beta.txt",
  CONCERTDR.compoundinfo_file = "/data/cmap/compoundinfo_beta.txt"
)

# 2. Filter siginfo to disease-relevant contexts
filtered_siginfo <- subset_siginfo_beta(
  getOption("CONCERTDR.siginfo_file"),
  interactive = FALSE,
  filters = list(
    pert_type  = "trt_cp",
    pert_itime = c("6 h", "24 h"),
    cell_iname = c("K562", "HL60", "THP1", "JURKAT")
  )
)

# 3. Prepare your disease signature
# Option A: read from a file
signature <- read.delim("my_disease_signature.txt")
# Option B: create from gene lists
# signature <- create_signature_from_gene_lists(
#   up_genes   = c("TP53", "MYC"),
#   down_genes = c("EGFR", "KRAS")
# )

# 4. Build reference matrix (this step may take a long time)
reference_df <- extract_cmap_data_from_siginfo(
  siginfo_file  = filtered_siginfo,
  geneinfo_file = getOption("CONCERTDR.geneinfo_file"),
  gctx_file     = getOption("CONCERTDR.gctx_file"),
  filter_quality = FALSE,
  landmark       = TRUE
)

# 5. Score
results <- process_signature_with_df(
  signature_file = signature,   # data.frame or file path
  reference_df   = reference_df,
  methods        = c("xsum", "ks"),
  topN           = 400,
  permutations   = 1000,
  save_files     = FALSE
)

# 6. Annotate
views <- annotate_drug_results(
  results_df     = results$results$xsum,
  sig_info_file  = filtered_siginfo,
  comp_info_file = getOption("CONCERTDR.compoundinfo_file"),
  output_dir     = "results",
  write_outputs  = TRUE
)

head(views$wetlab_drug_view)
```

# Session Info

```{r session-info}
sessionInfo()
```
