---
title: "Signature Matching with CONCERTDR"
author:
  - name: Hengxin Pan
    affiliation: Walter and Eliza Hall Institute of Medical Research
    email: pan.h@wehi.edu.au
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Signature Matching with CONCERTDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

# Overview

This vignette walks through the complete signature-matching pipeline:

1. Filter the CMap `siginfo_beta.txt` metadata to a relevant subset.
2. Prepare the query disease signature (from a file, data frame, or gene lists).
3. Build a gene-expression reference matrix from the GCTX file.
   *(This step can take a long time — the package will print a progress message.)*
4. Score and inspect the disease signature against the reference.

Step 1 requires the full CMap download and is shown with `eval=FALSE`.
Step 3 also requires the full CMap download and is shown with `eval=FALSE`.
Steps 2 and 4 use the bundled example data and are fully executable.

```{r load}
library(CONCERTDR)
```

# Step 1 — Filter Siginfo (Full CMap Data)

`subset_siginfo_beta()` lets you narrow the ~1.2 million CMap signatures to
the perturbation types and cell lines most relevant to your biological
question.

## Non-interactive filtering

The `filters` argument accepts a named list of column–value pairs. All
conditions are applied in sequence (logical AND).

```{r filter-noninteractive, eval=FALSE}
filtered_siginfo <- subset_siginfo_beta(
  siginfo_file = getOption("CONCERTDR.siginfo_file"),
  interactive  = FALSE,
  filters = list(
    pert_type  = c("trt_cp"),          # small-molecule compounds only
    pert_itime = c("6 h", "24 h"),     # standard time points
    cell_iname = c("HL60", "THP1",     # AML-relevant cell lines
                   "K562", "JURKAT",
                   "U937", "HAP1")
  )
)
cat("Retained signatures:", nrow(filtered_siginfo), "\n")
```

## Interactive filtering

Running with `interactive = TRUE` opens a console menu that shows, for each
filter dimension, exactly how many signatures each choice would retain along
with a preview of how many options the *next* dimension would have. This is
useful for exploratory data analysis.

```{r filter-interactive, eval=FALSE}
filtered_siginfo <- subset_siginfo_beta(
  siginfo_file = getOption("CONCERTDR.siginfo_file"),
  interactive  = TRUE,
  show_preview = TRUE
)
```

## Filterable columns

| Column | Typical values |
|--------|---------------|
| `pert_type` | `trt_cp` (compound), `trt_xpr` (CRISPR KO), `trt_sh` (shRNA), `trt_oe` (overexpression) |
| `pert_itime` | `6 h`, `24 h`, `96 h` |
| `pert_idose` | `0.04 uM`, `0.12 uM`, `0.37 uM`, `1.11 uM`, `3.33 uM`, `10 uM` |
| `cell_iname` | Over 70 cell lines; use `table(siginfo$cell_iname)` to browse |

## Demonstration with bundled siginfo

The bundled example siginfo has 10 signatures across five drugs.

```{r filter-demo}
example_siginfo_file <- system.file("extdata", "example_siginfo.txt",
                                     package = "CONCERTDR")
# Filter to K562 only
filtered_demo <- subset_siginfo_beta(
  siginfo_file = example_siginfo_file,
  interactive  = FALSE,
  filters      = list(cell_iname = "K562"),
  verbose      = TRUE,
  show_preview = FALSE
)
filtered_demo[, c("sig_id", "cmap_name", "cell_iname", "pert_itime", "pert_idose")]
```

# Step 2 — Prepare the Query Signature

The query signature must contain at minimum two columns:

- `Gene` — HGNC gene symbol (uppercase)
- `log2FC` — log2 fold-change between disease and control

Sort rows by `|log2FC|` descending so the highest-magnitude genes are
considered first when `max_genes` is applied.

## Loading a signature from a file

```{r load-signature}
sig_file  <- system.file("extdata", "example_signature.txt",
                          package = "CONCERTDR")
signature <- read.delim(sig_file)

# Separate up- and down-regulated genes
up_genes   <- signature$Gene[signature$log2FC > 0]
down_genes <- signature$Gene[signature$log2FC < 0]

cat("Up-regulated genes:  ", length(up_genes),   "\n")
cat("Down-regulated genes:", length(down_genes),  "\n")

# Preview the most-extreme genes
head(signature[order(-abs(signature$log2FC)), ], 8)
```

## Passing a signature as a data frame

You can pass a `data.frame` directly to `process_signature_with_df()` and
`extract_signature_zscores()` instead of a file path — this is convenient
when the signature is already in memory (e.g.\ from a differential-expression
analysis):
```{r sig-df-demo}
# 'signature' is already a data.frame from the previous chunk
head(signature)
# We will pass it directly in Step 4 below
```

## Creating a signature from gene lists

If you only have up- and down-regulated gene lists (without fold-changes),
`create_signature_from_gene_lists()` builds a minimal signature data frame
with default values of 1 and −1:

```{r gene-list-signature}
sig_from_lists <- create_signature_from_gene_lists(
  up_genes   = c("BRCA1", "TP53", "MYC"),
  down_genes = c("RB1", "PTEN", "APC")
)
sig_from_lists
```

# Step 3 — Build the Reference Matrix (Full CMap Data)

> **Note:** This step reads a large GCTX file and can take a long time
> (minutes to tens of minutes depending on the number of signatures and your
> hardware). The package will print a progress message — please do not
> interrupt the process.

`extract_cmap_data_from_siginfo()` reads the GCTX file for only the signature
IDs present in your filtered siginfo, returning a genes × signatures data
frame with siginfo metadata attached as an attribute.

```{r build-reference, eval=FALSE}
# Pass the in-memory filtered_siginfo directly — no need to write it to disk
reference_df <- extract_cmap_data_from_siginfo(
  siginfo_file   = filtered_siginfo,    # data.frame from Step 1
  geneinfo_file  = getOption("CONCERTDR.geneinfo_file"),
  gctx_file      = getOption("CONCERTDR.gctx_file"),
  filter_quality = FALSE,   # already filtered above
  landmark       = TRUE     # restrict to 978 landmark genes
)

cat("Reference matrix:", nrow(reference_df), "genes ×",
    ncol(reference_df) - 1, "signatures\n")

# Siginfo metadata is stored as an attribute
metadata <- attr(reference_df, "metadata")
head(metadata)
```

For the remainder of this vignette we use the pre-computed example:

```{r load-reference}
ref_file     <- system.file("extdata", "example_reference_df.csv",
                             package = "CONCERTDR")
reference_df <- read.csv(ref_file, row.names = 1, check.names = FALSE)
reference_df$gene_symbol <- rownames(reference_df)

cat("Reference:", nrow(reference_df), "genes ×",
    ncol(reference_df) - 1, "signatures\n")
```

# Step 4 — Run Signature Matching

## Running a single method

```{r run-single-method}
results_ks <- process_signature_with_df(
  signature_file = sig_file,
  reference_df   = reference_df,
  methods        = "ks",
  permutations   = 100,
  save_files     = FALSE
)
print(results_ks)
```

The top hits sorted by score (most negative = strongest reversal):

```{r top-hits-ks}
ks_df <- results_ks$results$ks
ks_df[order(ks_df$Score), ]
```

## Running multiple methods

```{r run-all-methods}
results_all <- process_signature_with_df(
  signature_file = sig_file,
  reference_df   = reference_df,
  methods        = c("ks", "xsum", "xcos", "zhang", "gsea0", "gsea1", "gsea2"),
  topN           = 4,          # for real data use topN = 400
  permutations   = 100,
  save_files     = FALSE
)
```

## Comparing methods side-by-side

```{r compare-methods}
# Extract the top-3 ranked compound per method
lapply(results_all$results, function(df) {
  head(df[order(df$Score), c("compound", "Score", "pAdjValue")], 3)
})
```

`summary()` aggregates across all methods with a combined weighted rank:

```{r summary-all-methods}
summary(results_all, top_n = 5)
```

## Score interpretation

```{r score-interpretation}
ks_df <- results_all$results$ks
ks_df$direction <- ifelse(
  ks_df$Score < 0, "Reversal (potentially therapeutic)",
  ifelse(ks_df$Score > 0, "Mimic / Aggravating", "Neutral")
)
ks_df[, c("compound", "Score", "direction")]
```

## Visualising scores

```{r plot-scores, fig.width=7, fig.height=4}
plot(results_all, method = "ks", plot_type = "scores", top_n = 10)
```

```{r plot-volcano, fig.width=6, fig.height=4}
plot(results_all, method = "xsum", plot_type = "volcano")
```

## Saving results to disk

Set `save_files = TRUE` and provide `output_dir` to write per-method CSV
files and a combined summary to disk.

```{r save-results, eval=FALSE}
results_saved <- process_signature_with_df(
  signature_file = sig_file,
  reference_df   = reference_df,
  methods        = c("ks", "xsum"),
  permutations   = 1000,
  output_dir     = "results",
  save_files     = TRUE     # writes sig_match_ks_results.csv, etc.
)
```

## Choosing `topN` and `permutations`

The `topN` parameter controls how many top- and bottom-ranked genes from each
reference profile are used by `xcos` and `xsum`. For the full 978 landmark
gene space, `topN = 400–500` is a common choice. Too small a value may miss
true signal; too large approaches the GSEA methods in behaviour.

The `permutations` parameter controls the resolution and stability of p-values.
Higher values give more reliable p-values but take longer to compute. For a
quick screen, `permutations = 100` is acceptable; for final results, use
`permutations = 1000` or more.

```{r topn-demo}
# Compare xsum at different topN values
scores_topn <- lapply(c(2, 4, 6), function(n) {
  res <- process_signature_with_df(
    signature_file = sig_file,
    reference_df   = reference_df,
    methods        = "xsum",
    topN           = n,
    permutations   = 50,
    save_files     = FALSE
  )
  df <- res$results$xsum[order(res$results$xsum$Score), ]
  df$topN <- n
  head(df, 3)
})
do.call(rbind, scores_topn)
```

# Session Info

```{r session-info}
sessionInfo()
```
