---
title: "Signature Matching with CONCERTDR"
author: "Hengxin Pan"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Signature Matching with CONCERTDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Signature Matching with CONCERTDR

This vignette demonstrates how to match a differential-expression signature
against CMap reference profiles and how to annotate and interpret the results.
All executable code uses the bundled example data. Sections that require the
full CMap download use `eval=FALSE`.

```{r setup}
library(CONCERTDR)
```

## Input Data

### 1. The Query Signature

A tab-delimited file with two columns:

- `Gene`: HGNC gene symbols
- `log2FC`: log2 fold-change (positive = up-regulated, negative = down-regulated)

```{r signature}
sig_file <- system.file("extdata", "example_signature.txt", package = "CONCERTDR")
sig <- read.delim(sig_file)
head(sig)
cat("Up-regulated genes:",   sum(sig$log2FC > 0), "\n")
cat("Down-regulated genes:", sum(sig$log2FC < 0), "\n")
```

### 2. The Reference Matrix

A data frame with genes as rows and CMap signature IDs as columns, produced
by `extract_cmap_data_from_config()` or `extract_cmap_data_from_siginfo()`
(see the *CMap Workflow* vignette). The bundled example is used here:

```{r reference}
ref_file <- system.file("extdata", "example_reference_df.csv", package = "CONCERTDR")
ref_df   <- read.csv(ref_file, row.names = 1, check.names = FALSE)
cat("Reference:", nrow(ref_df), "genes x", ncol(ref_df), "signatures\n")
ref_df[1:4, 1:4]
```

## Available Matching Methods

CONCERTDR wraps the [RCSM](https://github.com/Jasonlinchina/RCSM) package
and exposes these methods via the `methods` argument:

| Method | Argument | Description |
|--------|----------|-------------|
| Kolmogorov-Smirnov | `"ks"` | Non-parametric; direction-aware KS statistic |
| Extended cosine | `"xcos"` | Cosine similarity of ranked lists |
| Extended sum | `"xsum"` | Sum-based enrichment |
| GSEA weight 0 | `"gsea0"` | Unweighted GSEA |
| GSEA weight 1 | `"gsea1"` | Standard weighted GSEA |
| GSEA weight 2 | `"gsea2"` | Square-root weighted GSEA |
| Zhang score | `"zhang"` | Method from Zhang et al. 2008 |

## Running Signature Matching

`process_signature_with_df()` is the main function. It requires RCSM to be
installed. With the bundled example data:

```{r match, eval=FALSE}
# Requires: remotes::install_github("Jasonlinchina/RCSM")
results <- process_signature_with_df(
  signature_file = sig_file,
  reference_df   = ref_df,
  output_dir     = tempdir(),
  permutations   = 100,
  methods        = c("ks", "xcos", "xsum"),
  save_files     = FALSE
)

# results is a cmap_signature_result object
print(results)
summary(results)
```

### Inspecting Individual Method Results

```{r inspect-results, eval=FALSE}
# KS score data frame: columns compound, Score, pValue, pAdjValue, rank
ks_df <- results$results$ks
head(ks_df[order(ks_df$Score), ])          # Most negative = strongest reversal
```

### Interpreting Scores

- **Negative score** - the drug *reverses* the query signature (potentially therapeutic)
- **Positive score** - the drug *mimics* the query signature (possibly aggravating)
- **|Score| near 0** - no relationship

## Annotating Results with Drug Metadata

`annotate_drug_results()` merges scores with CMap compound metadata:

```{r annotate, eval=FALSE}
# Requires compoundinfo_beta.txt from clue.io
views <- annotate_drug_results(
  results_df    = ks_df,
  sig_info_file = "path/to/siginfo_beta.txt",
  comp_info_file = "path/to/compoundinfo_beta.txt",
  output_dir    = "results",
  write_outputs = TRUE
)

# Four analysis-ready tables
head(views$wetlab_drug_view)   # Drug-focused, de-duplicated
head(views$wetlab_gene_view)   # Gene-knockdown hits
head(views$tech_view_all)      # Full technical table
head(views$drug_context_summary) # Per-drug context count
```

## Fuzzy Drug Name Matching

`fuzzy_drug_match()` helps reconcile drug name variants using string
similarity (requires the `RecordLinkage` package):

```{r fuzzy, eval=FALSE}
# install.packages("RecordLinkage")
matches <- fuzzy_drug_match(
  query_names     = c("imatinb", "doxorubicn"),   # typos
  reference_names = c("imatinib", "dasatinib", "doxorubicin", "vincristine"),
  method          = "jaro",
  threshold       = 80
)
matches
```

## Visualising Results

When GCTX data is available, `plot_signature_direction_tile_barcode()` draws
a directional barcode heatmap showing which genes in the query signature are
up- or down-regulated across the top-scoring CMap signatures:

```{r visualise, eval=FALSE}
# Requires ComplexHeatmap, circlize, and the full CMap GCTX file
plot_signature_direction_tile_barcode(
  results_df     = ks_df,
  signature_file = sig_file,
  gctx_file      = "path/to/level5_beta_trt_cp_n720216x12328.gctx",
  geneinfo_file  = "path/to/geneinfo_beta.txt",
  siginfo_file   = "path/to/siginfo_beta.txt",
  max_genes      = 60,
  max_perts      = 30,
  save_png       = TRUE,
  output_png     = "barcode_heatmap.png"
)
```

## Complete One-Step Workflow

`run_cmap_workflow()` wraps extraction + matching into a single call:

```{r workflow, eval=FALSE}
all_results <- run_cmap_workflow(
  config_file   = "conf/my_config.conf",
  signature_file = sig_file,
  geneinfo_file = "path/to/geneinfo_beta.txt",
  siginfo_file  = "path/to/siginfo_beta.txt",
  gctx_file     = "path/to/level5_beta_trt_cp_n720216x12328.gctx",
  output_dir    = "results",
  methods       = c("ks", "xcos"),
  permutations  = 1000
)
```

## Session Info

```{r session-info}
sessionInfo()
```
